---
title: "4: Physical Properties of Lakes"
author: "Hydrologic Data Analysis | Kateri Salk"
date: "Fall 2019"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Lesson Objectives
1. Investigate the relationships between temperature, light, and oxygen in lakes by analyzing monitoring data
2. Analyze changes in lake surface temperatures over time
3. Apply data analytics skills to questions about physical properties of lakes
4. Communicate findings with peers through oral, visual, and written modes

## Opening Discussion
We analyzed temperature profiles and seasonal changes on Wednesday. What were some key takeaway points about thermal stratification and mixing?

<Epilimnion: surface layer. well mixed, Metalimnion: rapid change, middle. , Hypolimnion: cold bottom layer that separated from the surface layer layer by the meta...

We dont always have them. Really shallow lake that fully mixed. Spring and fall mixing.

Storng density gradient.

4°C max density liquid water. Preparing to become ice. Colder goes up.>

## Session Set Up
```{r, message = FALSE}
getwd()

library(tidyverse)
library(gridExtra)
library(cowplot)
library(lubridate) #lets you do things with dates.

NTLdata <- read.csv("./Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv")

theme_set(theme_classic())

```

## Data Wrangling and Exploration

### Investigate structure of dataset

```{r}
# Change sampledate to Date
NTLdata$sampledate <- as.Date(NTLdata$sampledate, "%m/%d/%y") # the format it comes in

# Make three data frames, one for each lake
Pauldata <- filter(NTLdata, lakename == "Paul Lake")
Peterdata <- filter(NTLdata, lakename == "Peter Lake")
Tuesdaydata <- filter(NTLdata, lakename == "Tuesday Lake")

# Make a data frame to inspect 2016 data for Paul Lake
Pauldata.2016 <- filter(Pauldata, year4 == 2016)

```

### Data Visualization

Let's look at profiles of temperature, irradiance, and dissolved oxygen over the course of a season. We will use the `plot_grid` function from the cowplot package to put the graphs together on one plot. Pay attention to the differences among the formats of each plot and how these work together in the final plot. 

```{r}
Tempprofiles2016 <- 
  ggplot(Pauldata.2016, aes(x = temperature_C, y = depth, color = daynum)) +
  geom_point() +
  scale_y_reverse() +
  scale_x_continuous(position = "top") +
  scale_color_viridis_c(end = 0.8, option = "magma") + #not use the yellow (you can specify end and start colors)
  labs(x = expression("Temperature "(degree*C)), y = "Depth (m)") +
  theme(legend.position = "none")
print(Tempprofiles2016)

Lightprofiles2016 <- 
  ggplot(Pauldata.2016, aes(x = irradianceWater, y = depth, color = daynum)) +
  geom_point() +
  scale_y_reverse() +
  scale_x_continuous(position = "top") +
  scale_color_viridis_c(end = 0.8, option = "magma") +
  labs(x = expression("Irradiance (W m"^-2*")"), y = "Depth (m)") +
  theme(legend.position = "none", 
        axis.text.y = element_blank(), axis.title.y = element_blank())
print(Lightprofiles2016)

Oxygenprofiles2016 <- 
  ggplot(Pauldata.2016, aes(x = dissolvedOxygen, y = depth, color = daynum)) +
  geom_point() +
  scale_y_reverse() +
  scale_x_continuous(position = "top") +
  scale_color_viridis_c(end = 0.8, option = "magma") +
  labs(x = expression("DO (mg L"^-1*")"), y = "Depth (m)", 
       color = "Julian Day") +
  theme(axis.text.y = element_blank(), axis.title.y = element_blank())
print(Oxygenprofiles2016)

Physicalprofiles2016 <- 
  plot_grid(Tempprofiles2016, Lightprofiles2016, Oxygenprofiles2016, #int he cow package
            ncol = 3, rel_widths = c(1.25, 1, 1.5))
print(Physicalprofiles2016)

#she kept the ticks in the y axis. To keep reference.
```

How do temperature profiles vary over the course of a season? What are the mechanisms for this pattern?

> It always goes down with depth but the range is higher during the summer. Temperatures in the hypolimnion dont change. Air temperture influence the temperature in the epilimnion. Ther eis mixing there. Meta is a transition to the hypo where there is no effect.

How do irradiance profiles vary over the course of a season? What are the mechanisms for this pattern?

> The irradience goes down with depth. Is higher in summer. There is less effect in shallow waters. You could have algae blooms or cloudy day. (explaining the orange minimum)

How do dissolved oxygen profiles vary over the course of a season? What are the mechanisms for this pattern?

> Is constant upto 2.5 meters then it goes up and then sudden drop. DO goes up and temperature goes down. MOre T less DO.

>Biologicaly and chemicaly active.
>Colder water holds more water. 
> Goes up to saturation because it drops temperature and photosynthesis. the phytoplancton hangs up in the top of the hypolimniun. Enough light and nutrients.
>Hypolimnion. Respiration (bacteria, zooplancton). Phytoplancton dies, goes down and bacteria creates Nutrients. It is sudden because the phytoplancton is the source and cant be that deep.

## Data Analysis

### Long-term change in temperature
How much have surface temperatures increased over the period of study? Is this increase significant? Isolate surface depths and run the test for May, June, July, and August temperatures separately. Use a linear regression with year as the predictor variable and temperature as the response variable.

Steps: 

1. Add a column named "Month" to the data frame (hint: lubridate package)
2. Filter your data frame so that it only contains surface depths and months 5-8
3. Create 4 separate data frames, one for each month
4. Run a linear regression for each data frame (see instructions above)
5. For significant trends, calculate how many degrees the lake has warmed over the period of study
5. Create a ggplot of surface temps by date and facet by month. Add `geom_smooth(se = FALSE, method = lm)` to the plot to display the regression line.

```{r}
Pauldata$Month <- month(Pauldata$sampledate)
Pauldata.Winter <- filter(Pauldata, Month == 5 | Month == 6 | Month == 7 | Month == 8)

Pauldata.May <- filter(Pauldata, Month == 5)
Pauldata.June <- filter(Pauldata, Month == 6)
Pauldata.July <- filter(Pauldata, Month == 7)
Pauldata.August <- filter(Pauldata, Month == 8)

lm.may <- lm(Pauldata.May$temperature_C~Pauldata.May$year4)
summary(lm.May)

lm.June <- lm(Pauldata.June$temperature_C~Pauldata.June$year4)
summary(lm.June)

lm.July <- lm(Pauldata.July$temperature_C~Pauldata.July$year4)
summary(lm.July)

lm.August <- lm(Pauldata.August$temperature_C~Pauldata.August$year4)
summary(lm.August)

Tempchange.plot <-
  ggplot(Pauldata.Winter, aes(x = sampledate, y = temperature_C)) +
  geom_point() +
  geom_smooth(se=FALSE, method =lm) +
  facet_grid(rows=vars(Month))
  print(Tempchange.plot)

# Think aboutthe units. Is it importnat 1°C.

```

Summarize your results below. How would you explain your findings and the importance of these findings to a non-specialist audience while still employing the data and statistical results?

>

## Closing Discussion
How do warming surface waters impact the physical properties of a lake? How might this affect the ecology of the lake?

How do we reconcile large amounts of variability in long-term records?
